Non-word repetition in children learning Yélî Dnye

### 2021-02-11 manuscript.tex
proofing

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[english,,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Non-word repetition in Yélî Dnye},
            pdfkeywords={phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{NWR in Yélî Dnye}
\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
\authornote{

Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~9,000 words}
\usepackage{lineno}

@@

\title{Non-word repetition in Yélî Dnye}

@@

\abstract{
In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation in language development at a global and individual level via three contributions. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order. Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
}

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014). NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007). Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas.

The first research area is at the crossing of typology and phonological development. Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely linguistic skill. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly jittered for ease of inspection of different lengths at a given age.}
\end{figure}

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021). In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance et al., 2005). Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007). Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on these too.

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size). Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities Maddieson, 2013b, and @maddieson2013consonant with no exceptions to our knowledge). There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, Cristia et al. (2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991). Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills. Indeed, Cristia et al. (2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from Cristia et al. (2020)'s hypotheses just mentioned. Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.

Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system (Levinson, 2020), is unlike any other in the region. In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.). Thus, with respect to our first research goal, Yélî Dnye is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.

@@

Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020). The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} /ṇṭæp/ \emph{Spondylus shell}) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \enquote{ndapî} /'ṇṭæ.pɯ/). There are also a handful of words starting with /æ/ (e.g., \enquote{ala} /æ.'læ/ \emph{here}) and a small collection of single-vowel grammatical morphemes (see Levinson (2020) for details).

@@

Most Yélî parents are swidden horticulturalists. Within a group of households, it is often the case that older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them less time to spend directly with the children in their household than may be the case in infants reared by stay-at-home parents in urban sites. Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020). Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available (Casillas et al., 2020). Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013). We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 40 final items distributed across children.

@@

A second list of candidate items was generated in a second trip to the island in 2019, when data were collected, by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded. Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, 2020), it was also possible that we initially generated items with illegal, but currently undocumented constraints. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were then presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded, based on their orthographic forms, using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) by the same informant, monitored by the second author for clear production of the phonological target. The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present. Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \enquote{yî} /yɯ/; identified as \enquote{yi} /yi/, \emph{tree}). Additionally, an error was made when preparing files for annotation, resulting in two items being merged (\enquote{tpâ} /tpɑ/ and \enquote{tp:a} /tpæ̃/). These three problematic items are not described here, and removed from the analyses below.

The final list includes three practice items and 40 test items (across infants): 16 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).

@@

A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The 40 elicitation sets are available online from \href{https://osf.io/dtxue/}{osf.io/dtxue/}. The split had the following constraints:

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[english,,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Non-word repetition in children learning Yélî Dnye},
            pdfkeywords={phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{NWR in Yélî Dnye learners}
\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
\authornote{

Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~11,500 words}
\usepackage{lineno}

@@

\title{Non-word repetition in children learning Yélî Dnye}

@@

\abstract{
In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
}

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are ({\textbf{???}}), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014). NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007). Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016). NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas, three via direct results.

The first research area is at the crossing of typology and phonological development. Previous work using NWR has preferred relatively universal and early-acquired phonemes (with exceptions including Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely linguistic skills. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; {\textbf{???}}), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese ({\textbf{???}}); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly shifted for ease of inspection of different lengths at a given age.}
\end{figure}

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based ({\textbf{???}}). In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age ({\textbf{???}}; Farmani et al., 2018; Vance et al., 2005). Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; {\textbf{???}}; Balladares, Marshall, \& Griffiths, 2016; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007). Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size). Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b); Maddieson (2013a); with no exceptions to our knowledge{]}. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, Cristia et al. (2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance (e.g., Gathercole, Willis, \& Baddeley, 1991). Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills. Indeed, Cristia et al. (2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. Aside from Cristia et al. (2020)'s hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations). Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.

Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system (Levinson, 2020), is unlike any other in the region. In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.). Thus, with respect to our first research goal, Yélî Dnye is a good language to test because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.

@@

Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020). The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} /ṇṭæp/ \emph{Spondylus shell}) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \enquote{ndapî} /'ṇṭæ.pɯ/). There are also a handful of words starting with /æ/ (e.g., \enquote{ala} /æ.'læ/ \emph{here}) and a small collection of single-vowel grammatical morphemes (see Levinson (2020) for details).

@@

Most Yélî parents are swidden horticulturalists. Within a group of households, it is often the case that older adolescents and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities. Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020). Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available (Casillas et al., 2020). Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013). We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; {\textbf{???}}), with the entire test inventory of 40 final items distributed across children.

@@

A second list of candidate items was generated in a second trip to the island in 2019, when data were collected, by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded. Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, 2020), it was also possible that we initially generated items with illegal, but currently undocumented constraints. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence). These candidates were then presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded, based on their orthographic forms, using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) by the same informant, monitored by the second author for clear production of the phonological target. The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present. Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \enquote{yî} /yɯ/; identified as \enquote{yi} /yi/, \emph{tree}). Additionally, an error was made when preparing files for annotation, resulting in two items being merged (\enquote{tpâ} /tpɑ/ and \enquote{tp:a} /tpæ̃/). These three problematic items are not described here, and removed from the analyses below.

The final list includes three practice items and 40 test items (across infants): 16 monosyllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).

@@

A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sub-lists, to generate 40 different elicitation sets. The 40 elicitation sets are available online from \href{https://osf.io/dtxue/}{osf.io/dtxue/}. The split had the following constraints:
